Affective Uplift During Video Game Play: A Naturalistic Case Study

### 2024-02-28 ms.Rmd
MS revised after reviwe

---
title: "Affective Uplift During Video Game Play: A Naturalistic Case Study"
shorttitle: Affective Uplift During Play
leftheader: Affective Uplift During Play
author: 
  - name: Matti Vuorre
    corresponding: yes
    affiliation: "1,2*"
    address: Tilburg University
    email: mjvuorre@uvt.nl
  - name: Nick Ballou
    affiliation: "2*"
  - name: Thomas Hakman
    affiliation: "2"
  - name: Kristoffer Magnusson
    affiliation: "2,3"
  - name: Andrew K. Przybylski
    affiliation: "2"
    address: Oxford Internet Institute
    email: andy.przybylski@oii.ox.ac.uk
affiliation:
  - id: 1
    institution: Tilburg School of Social and Behavioral Sciences, Tilburg University
  - id: 2
    institution: Oxford Internet Institute, University of Oxford
  - id: 3
    institution: Centre for Psychiatry Research, Department of Clinical Neuroscience, Karolinska Institutet, & Stockholm Health Care Services, Region Stockholm, Sweden
authornote: |
  \noindent \textbf{This pre-print is not yet peer-reviewed.}

  *MV and NB contributed equally to this work.

wordcount: "4,802"
bibliography: references.bib
keywords: "video games, well-being, mood, play behavior, telemetry"
csl: apa.csl
floatsintext: yes
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_docx: 
    number_sections: false
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: true
editor_options: 
  chunk_output_type: console
---

@@

# Document options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.asp = 0.618
)

# mcmc settings
MAX_CORES <- as.numeric(Sys.getenv("N_CORES"), 4)
N_ITER <- as.numeric(Sys.getenv("N_ITER"), 2000)
options(
  brms.backend = Sys.getenv("BRMS_BACKEND", "rstan"), 
  brms.threads = as.numeric(Sys.getenv("N_THREADS"), 1),
  mc.cores = MAX_CORES
)
```

@@

Understanding whether and when games' short-term effects emerge is critical for establishing games' potential for mood-related interventions, as well as for building a theoretical foundation for repeated short-term gaming experiences' long-term effects on mental health. At present, however, the prevalence and magnitude of these short-term effects remains poorly understood. Some existing evidence suggests that games can provide short-term boosts to well-being ["mood repair", @bowmanMoodGameSelective2015; @tyackRestorativePlayVideogames2020], possibly to a greater extent than non-interactive media such as videos [@riegerEatingGhostsUnderlying2015]. On the other hand, games might affect players negatively: Frustrating gaming experiences, for example, can lead to negative consequences such as immediate post-play aggression [@przybylskiCompetenceimpedingElectronicGames2014].

Despite the above examples, the validity and generalizability of research on games' short-term affective effects remains limited by three challenges. First, a substantial portion of gameplay research has relied on artificial stimuli; games created or substantially modified by academic researchers. While such customized games allow for greater experimental control, they are unlikely to reflect actual games' rich complexity [@mcmahanConsiderationsUseCommercial2011]. This issue of the limited ecological validity and generalizability of research stimuli (games) limits current inferences about popularly played games' psychological effects.

The second challenge is providing an ecologically valid *context* for play. Research participants typically play games in (online or physical) labs that do not resemble the natural contexts of play, such as when, with whom, and why people choose to play. In lab settings, research participants play to satisfy study requirements, rather than the intrinsic motivations that typically lead them to play. While beneficial to clarifying causal inferences, the extrinsically motivated play behaviors necessary in lab studies might relate differently to well-being than intrinsically motivated naturally occurring play [@bruhlmannMotivationalProfilingLeague2020; @howardStudentMotivationAssociated2021]. Therefore, results from such studies are less likely to accurately generalize to how games are played in the real world.

@@

In this study, we analyzed data from a large open dataset on PowerWash Simulator (PWS) play and psychological experiences [@vuorreIntensiveLongitudinalDataset2023]. The data was collected in a research edition of PWS that recorded gameplay events, game status records, and responses to psychological survey items. We developed the research edition of PWS in collaboration with PWS's developer, FuturLab, who made it freely available on Steam to anyone who owned the original game (£19.99 on 2023-09-20). From the players' perspective, the research edition was nearly identical to that of the main game with the addition of in-game pop-ups that inquired about psychological states during play.

@@

Although the overall magnitude of the estimated change was small (`r filter(fit_tbl, variable=="Difference")$res`), it nevertheless is comparable to mood shifts following other frequent voluntary activities. For example, experience sampling studies of US [@killingsworthWanderingMindUnhappy2010] and Korean [@choiTakingStockHappiness2017] adults indicated that watching television (+2% change compared to an individual's average across all activities), reading (+2%), and shopping (+3%) were associated with smaller mood shifts than we observed in association with PWS. On the other hand, those studies indicated that listening to music (+7–9%), eating or cooking (+8%), taking a walk (+9%), visiting an urban green space [+9–10%, @tostNeuralCorrelatesIndividual2019], exercising (+11%), dating (26.5%), sexual intercourse (+28%), or taking a trip (+30.5%) correlate with greater shifts in well-being.

Another interpretation of our result is that although the estimated PWS-play associated mood uplift is small, it is potentially large enough to be subjectively perceived. @anvariUsingAnchorbasedMethods2021 suggest that, on average, people are able to subjectively perceive a change of 2% in well-being on the related PANAS scale. Since our estimate of `r filter(fit_tbl, variable=="Difference")$res` is greater, we tentatively suggest that gaming, on average, is associated with mood uplifts that are large enough to be consciously felt by players.

@@

@@

Current theories provide a few candidates regarding mood change mechanisms. Mood management theory [MMT, @zillmannMoodManagementCommunication1988] posits that players choose specific entertainment media as part of a (conscious or subconscious) strategy to regulate valence and arousal and to increase hedonic well-being. Specifically, MMT predicts that when players feel more negative emotions, they are more likely to select games with positive hedonic valence (i.e., games that are generally positive in tone), and that when players are in high arousal states such as anger stress, they are more likely to choose calm and relaxing games with.

PowerWash Simulator provides an ideal case study for mood management theory, in that it is typically described as a low cognitive effort game: Players are known to play semi-idly with only partial focus, or even while consuming other media. As one reviewer writes, "When my brain is tired, or I'm not in the mood to compete or struggle in any way, I grab a power washer" [@marshallPowerWashSimulatorMy2021]. Accordingly, PWS may be especially well-suited for players who are emotionally or cognitively depleted, such as those with persistently low mood [@helsbyBandwidthComesGoes2023]---something that is partially supported by our findings that sessions with lower initial moods were those with larger uplifts during play. Although outside the scope of this study, we believe that the additional telemetry in the PWS dataset [@vuorreIntensiveLongitudinalDataset2023] would be valuable in addressing more nuanced questions regarding MMT.

Another candidate derives from self-determination theory [@ryanSelfdeterminationTheoryBasic2017]. At least one laboratory study has shown that the experience of basic psychological need satisfaction during media consumption moderates the change from pre–post play well-being [@tyackRestorativePlayVideogames2020]. Another found that need satisfaction in two commercial games, Animal Crossing: New Horizons and Plants vs Zombies: Battle for Neighborville, was independently associated with well-being, regardless of the amount of playtime [@johannesVideoGamePlay2021]. It may be the case, then, that experiences with greater feelings of mastery and effectiveness (competence satisfaction) or control and volition (autonomy satisfaction) would result in larger mood increases during PWS play. While we did not test these causal mechanisms between play and well-being in the current study, we anticipate several theory-driven detailed investigations of the PWS dataset [@vuorreIntensiveLongitudinalDataset2023] that might do so.

Second, we do not know whether the mood uplift associated with PWS play lingers or disappears after the gaming session ends. Answering this question would afford a better understanding of the motivations behind play: To what extent do people play games for their immediate hedonic benefits in contrast to the emotional and cognitive state players find themselves in after stopping? The difference between these two experiences might be an important differentiator between, for example, maladaptive *escaping from* and adaptive *escaping to* [@stensengAreThereTwo2021]. To answer this question, however, in-game telemetry data will need to be complemented with out-of-game psychological measures assessing well-being in the period after play. We anticipate that experience sampling methods, which are common in social media research [@aalbersCaughtMomentAre2022; @verbeijExperienceSamplingSelfreports2022] but have yet to make meaningful inroads in video games research, would be particularly valuable in addressing this question.

>>>>

---
title: "Affective Uplift During Video Game Play: A Naturalistic Case Study"
shorttitle: Affective Uplift During Play
leftheader: Affective Uplift During Play
author: 
  - name: Matti Vuorre
    corresponding: yes
    affiliation: "1,2"
    address: Tilburg University
    email: mjvuorre@uvt.nl
  - name: Nick Ballou
    affiliation: "2"
  - name: Thomas Hakman
    affiliation: "2"
  - name: Kristoffer Magnusson
    affiliation: "2,3"
  - name: Andrew K. Przybylski
    affiliation: "2"
    address: Oxford Internet Institute
    email: andy.przybylski@oii.ox.ac.uk
affiliation:
  - id: 1
    institution: Tilburg School of Social and Behavioral Sciences, Tilburg University
  - id: 2
    institution: Oxford Internet Institute, University of Oxford
  - id: 3
    institution: Centre for Psychiatry Research, Department of Clinical Neuroscience, Karolinska Institutet, & Stockholm Health Care Services, Region Stockholm, Sweden
authornote: |
  \noindent \textbf{This pre-print is not yet peer-reviewed.}

wordcount: "4,802"
bibliography: references.bib
keywords: "video games, well-being, mood, play behavior, telemetry"
csl: apa.csl
floatsintext: yes
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: true
  papaja::apa6_docx: 
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

# Document options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.asp = 0.618
)
```

@@

Understanding whether and when games' short-term effects emerge is critical for establishing games' potential for mood-related interventions, as well as for building a theoretical foundation for repeated short-term gaming experiences' long-term effects on mental health. Substantial existing evidence suggests that games can provide short-term boosts to well-being [@bowmanMoodGameSelective2015; @tyackRestorativePlayVideogames2020], possibly to a greater extent than non-interactive media such as videos [@riegerEatingGhostsUnderlying2015]. Much of that work took place under the "mood repair" and "mood management" labels [@zillmannMoodManagementCommunication1988], which describe how media might support users in balancing internal states following unpleasant feelings, possibly through addressing basic psychological needs [@BallouDeterding2023Basic; @ReineckeEtAl2012characterizing; @riegerEatingGhostsUnderlying2015; @TamboriniEtAl2011Media; @Tyack2019Need]. On the other hand, games might also affect players negatively: Frustrating gaming experiences, for example, can lead to negative consequences such as immediate post-play aggression [@przybylskiCompetenceimpedingElectronicGames2014].

At present, however, the prevalence and magnitude of these short-term effects remain poorly understood. Despite the above examples, the validity and generalizability of research on games' short-term affective effects remains limited by three challenges. First, a substantial portion of gameplay research has relied on artificial stimuli; games created or substantially modified by academic researchers [@bowmanMoodGameSelective2015]. While such customized games allow for greater experimental control, they are unlikely to reflect actual games' rich complexity [@mcmahanConsiderationsUseCommercial2011]. This issue of the limited ecological validity and generalizability of research stimuli (games) limits current inferences about popularly played games' psychological effects.

The second challenge is providing an ecologically valid *context* for play. Research participants typically play games in (online or physical) labs that do not resemble the natural contexts of play, such as when, with whom, and why people choose to play [@tyackRestorativePlayVideogames2020]. In lab settings, research participants play to satisfy study requirements, rather than the intrinsic motivations that typically lead them to play. While beneficial to clarifying causal inferences, the extrinsically motivated play behaviors necessary in lab studies might relate differently to well-being than intrinsically motivated naturally occurring play [@bruhlmannMotivationalProfilingLeague2020; @howardStudentMotivationAssociated2021]. Therefore, results from such studies are less likely to accurately generalize to how games are played in the real world.

@@

In this study, we analyzed data from a large open dataset on PowerWash Simulator (PWS) play and psychological experiences [@vuorreIntensiveLongitudinalDataset2023]. The data was collected in a research edition of PWS that recorded gameplay events, game status records, participant demographics, and responses to psychological survey items. We developed the research edition of PWS in collaboration with PWS's developer, FuturLab, who made it freely available on Steam to anyone who owned the original game (£19.99 on 2023-09-20). From the players' perspective, the research edition was nearly identical to that of the main game with the addition of in-game pop-ups that inquired about psychological states during play.

@@

Although the overall magnitude of the estimated change was small considering the scale range (`r filter(fit_tbl, variable=="Difference")$res`), comparative evidence from other frequent voluntary activities indicates that it might still be of meaningful magnitude. For example, experience sampling studies of US [@killingsworthWanderingMindUnhappy2010] and Korean [@choiTakingStockHappiness2017] adults indicated that watching television (+2% change compared to an individual's average across all activities), reading (+2%), and shopping (+3%) were associated with smaller mood shifts than we observed in association with PWS. On the other hand, those studies indicated that listening to music (+7–9%), eating or cooking (+8%), taking a walk (+9%), visiting an urban green space [+9–10%, @tostNeuralCorrelatesIndividual2019], exercising (+11%), dating (26.5%), sexual intercourse (+28%), or taking a trip (+30.5%) correlate with greater shifts in well-being.

Another interpretation of our result is that although the estimated PWS-play associated mood uplift is small, it is potentially large enough to be subjectively perceived. @anvariUsingAnchorbasedMethods2021 suggest that, on average, people are able to subjectively perceive a change of 2% in well-being on the related PANAS scale. Since our estimate of `r filter(fit_tbl, variable=="Difference")$res` is greater, we tentatively suggest that gaming, on average, is associated with mood uplifts that are large enough to be consciously experienced by players. This might hold especially for people who started the session with lower pre-play mood (Figure \@ref(fig:fig-ct2)), a finding that aligns with literature indicating that people in particularly low moods will selectively seek out media with potential to uplift and balance mood [@ReineckeEtAl2012characterizing].

@@

We believe the mood uplift observed here is significant for three reasons: First, it compares in magnitude to associations with other activities commonly considered as mood-enhancing. Second, the uplift is greater for individuals who started with a lower mood. Third, anticipatory effects, whereby mood increases before play in anticipation of this rewarding activity, are likely and would negatively bias the uplift observed here as an estimate of the overall effect. Nevertheless, we aimed to provide information about plausible mood-uplift magnitudes, whose interpretation likely depends on their context in future studies and applications.

@@

Current theories provide a few candidates regarding mood change mechanisms. Mood management theory [MMT, @zillmannMoodManagementCommunication1988] posits that people choose media partly as strategies to regulate valence and arousal and to increase hedonic well-being. Specifically, MMT predicts that when players feel more negative emotions, they are more likely to select games with positive hedonic valence (i.e., games that are generally positive in tone), and that when players are in high arousal states such as anger stress, they are more likely to choose calm and relaxing games with.

PowerWash Simulator provides an ideal case study for mood management theory, in that it is typically described as a low cognitive effort game: Players are known to play semi-idly with only partial focus, or even while consuming other media. As one reviewer writes, "When my brain is tired, or I'm not in the mood to compete or struggle in any way, I grab a power washer" [@marshallPowerWashSimulatorMy2021]. Accordingly, PWS may be especially well-suited for players who are emotionally or cognitively depleted, such as those with persistently low mood [@helsbyBandwidthComesGoes2023]---something that is partially supported by our findings that sessions with lower initial moods were those with larger uplifts during play. We think the consistency of this uplift---with over 70% of people predicted to experience uplifts---means that PWS and similar games might be especially effective for stabilizing noxious moods rather than enhancing neutral ones. Although outside the scope of this study, we believe that the additional telemetry in the PWS dataset [@vuorreIntensiveLongitudinalDataset2023] would be valuable in addressing more nuanced questions regarding MMT.

Another candidate derives from Self-determination theory [@ryanSelfdeterminationTheoryBasic2017]. At least one laboratory study has shown that the experience of basic psychological need satisfaction during media consumption moderates the change from pre–post play well-being [@tyackRestorativePlayVideogames2020]. Another found that need satisfaction in two commercial games, Animal Crossing: New Horizons and Plants vs Zombies: Battle for Neighborville, was independently associated with well-being, regardless of the amount of playtime [@johannesVideoGamePlay2021]. It may be the case, then, that experiences with greater feelings of mastery and effectiveness (competence satisfaction) or control and volition (autonomy satisfaction) would result in larger mood increases during PWS play. While we did not test these causal mechanisms between play and well-being in the current study, we anticipate several theory-driven detailed investigations of the PWS dataset [@vuorreIntensiveLongitudinalDataset2023] that might do so.

Second, we do not know whether the mood uplift associated with PWS play lingers or disappears after the gaming session ends. PWS might function not only as a mood-enhancer, but also as a stabilizer. Considering the generally high pre-play moods observed here, when compared to previous studies [@choiTakingStockHappiness2017; @killingsworthWanderingMindUnhappy2010], PWS player might indeed aim to sustain rather than reach an above-average mood. Of course, data on mood before, during, and after play is needed to answer this question, whose answer would afford a better understanding of the motivations behind play: To what extent do people play games for their immediate hedonic benefits in contrast to the emotional and cognitive state players find themselves in after stopping? The difference between these two experiences might be an important differentiator between, for example, maladaptive *escaping from* and adaptive *escaping to* [@stensengAreThereTwo2021]. To answer this question, however, in-game telemetry data will need to be complemented with out-of-game psychological measures assessing well-being in the period after play. We anticipate that experience sampling methods, which are common in social media research [@aalbersCaughtMomentAre2022; @verbeijExperienceSamplingSelfreports2022] but have yet to make meaningful inroads in video games research, would be particularly valuable in addressing this question.
